% !TeX root = main.tex
\lecture{15}{Tue 10 Mar 2026 12:00}{Statistical Physics II: The Atmosphere}

\section{The Atmosphere}
The very outer part of the atmosphere, from $90$km to $350$km is the ionosphere. This acts as a mirror which reflects EM radiation 

% TODO some waffle.

\subsection{Isothermal Model of the Atmosphere}
Consider an observer sitting in the troposphere above the earth, looking at a small cuboidal slice of the troposphere.

We will assume that:
\begin{itemize}
    \item The atmosphere is isothermal. In reality, increasing height reduces temperature by approx. $10$K per km.
    \item The atmosphere is solely comprised of an ideal gas. At the low pressure scales relevant for the atmosphere, this is a reasonable assumption.
    \item The earth is flat\footnote{Oh dear\dots}. Technically, assume that we we consider small enough heights and cross-sectional areas that we can disregard the curvature of the earth.
    \item The atmosphere is stationary and is therefore in mechanical equilibrium. This can be a good approximation depending on weather conditions.
\end{itemize}

The slab of atmosphere we're considering has area $A$, height $dh$ and sits at height $h$ above the surface of the earth. From the ideal gas state equation:
\[
    PV = nRT = N k_B T
\]
\[
    \implies P = \frac{N}{V} k_B T
\]
\[
    P = \rho_N k_B T
\]

Where $\rho_N$ is number density. %TODO.

The slab of atmosphere has two forces acting upon it, a force due to pressure and a force due to gravity. As we assume that slice is in mechanical equilibrium, we can equate these forces.

At the top of the slab, force due to gravity has a larger magnitude than the bottom, hence force due to pressure must also be larger at the top of the slab. As we consider a slice of infinitesimal height, we need not integrate and can just consider both values.

This gives us:
\[
    \text{force due to gravity } = \text{ force due to pressure}
\]
\[
    mg \rho_N V = A \rho(h) - A \rho(h + dh)
\]
\[
    mg \rho_N V = -d \rho(h) A
\]
\[
    mg \rho_N \times dh \times A = - d \rho(h) A
\]
\[
    mg \rho_N \times dh =  -d \rho(h)
\]
\[
    \frac{d \rho(h)}{dh} - mg \rho(N)
\]
\[
    \int d \rho(h) = - \int mg \rho_N dh
\]

And considering number density in the density integral, as %TODO
\[
    \int_{\rho(0)}^{\rho(h)} \frac{1}{\rho_N} d \rho(h) = - \int_{0}^{h} mg dh
\]

And substituting $dp = d \rho_N k_B T$ to give a LHS integral entirely in number density:
\[
    \int_{\rho_N(0)}^{\rho_N(h)} \frac{k_B T}{\rho_N} d \rho_N = -mgh    
\]
\[
    \int_{\rho_N(0)}^{\rho_N(h)} \frac{k_B T}{\rho_N} d \rho_N = -\frac{mgh}{k_B T}   
\]
\[
    \left[\ln \left(\rho_N\right)\right]^{\rho_N(h)}_{\rho_N(0)} = -\frac{mgh}{k_B T}   
\]
\[
    \frac{\rho_N(h)}{\rho_N(0)} = \exp \left(- \frac{mgh}{k_B T}\right) \implies \boxed{\rho_N(h) = \rho_N(0) \exp \left(- \frac{mgh}{k_B T}\right)}
\]










